Fast and scalable Gaussian process modeling with applications to astronomical time series \vspace{-3\baselineskip}

### 2017-03-21 paper/ms.tex
removing some todos
finished paragraph on power spectrum

\begin{figure}[!htbp]
\begin{center}
\includegraphics[width=0.5\textwidth]{figures/rotation-period.pdf}
\caption{The posterior constraint on the rotation period of KIC~1430163 using
    the dataset and model from \Figure{rotation}.
    The period is the parameter $P$ in \eq{rot-kernel} and this figure shows
    the posterior distribution marginalized over all other nuisance parameters
    in \eq{rot-kernel}.
    This is consistent with the published rotation period made using the
    autocorrelation function and the full \kepler\ data (\dfmtodo{CITE}).
    \figurelabel{rotation-period}}
\end{center}
\end{figure}

@@

Our background is in studying transiting exoplanets, a field which has only
recently begun to adopt full covariance matrices in analyzing the noise in
stellar lightcurves when detecting or characterizing transiting planets
\citep[for example,][]{Gibson:2012, Barclay:2015, Evans:2015, Aigrain:2016,
Grunblatt:2016, Luger:2016}.
One promising early attempt at accounting for correlated noise was the wavelet
approach of \citet{Carter:2009}.
Their technique allowed for a power-law spectrum for the noise scaling as
$\omega^{-\alpha}$, which unfortunately is not a normalizable, nor
physically-accurate, description of stellar variability.
Their approach runs in ${\cal O}(N)$ time for $\alpha=1$, which is referred to
as `flicker', $1/r$, or pink noise;  however, this particular type of noise
overpredicts the power spectrum at low frequency compared to the Solar
granulation power spectrum.  It also cannot describe quasi-periodic noise.

@@

The \celerite\ formalism also allows a principled approach to the estimation
of the power spectrum of a time series, as well as its uncertainties.
Fourier transforms work for uniformly sampled data;  astronomical data are
rarely (if ever) uniformly sampled.  Periodograms are affected by the window
function and aliasing in datasets, and make estimating the significance of
spectral features ambiguous.  The Bayesian approach to the estimation of a power
spectrum (CITE) avoids these difficulties:  it allows the computation of
uncertainties, it accounts for the window function on the uncertainties of
the power spectrum, and in principle it can treat
\agoltodo{Finish this\ldots}

@@

For arbitrary $J$, a procedure based upon Sturm's theorem \citep{Dorrie:1965}
allows one to determine whether there are any real roots within the range
$(0,\,\infty]$.
We first construct $Q_0(z)$ and it's derivative $Q_1(z) = {Q_0}^\prime(z)$,
and then loop from $k=2$ to $k=2\,(J-1)+1$, computing
\begin{eqnarray}
Q_k(z) = -{\rm rem}(Q_{k-2},\,Q_{k-1})
\end{eqnarray}
where the function ${\rm rem}(p,\,q)$ is the remainder polynomial after
dividing $p(z)$ by $q(z)$.

>>>>

\begin{figure}[!htbp]
\begin{center}
\includegraphics[width=0.5\textwidth]{figures/rotation-period.pdf}
\caption{The posterior constraint on the rotation period of KIC~1430163 using
    the dataset and model from \Figure{rotation}.
    The period is the parameter $P$ in \eq{rot-kernel} and this figure shows
    the posterior distribution marginalized over all other nuisance parameters
    in \eq{rot-kernel}.
    This is consistent with the published rotation period made using the
    autocorrelation function and the full \kepler\ baseline
    \citep{Mathur:2014}.
    \figurelabel{rotation-period}}
\end{center}
\end{figure}

@@

Our background is in studying transiting exoplanets, a field which has only
recently begun to adopt full covariance matrices in analyzing the noise in
stellar lightcurves when detecting or characterizing transiting planets
\citep[for example,][]{Gibson:2012, Barclay:2015, Evans:2015, Aigrain:2016,
Grunblatt:2016, Luger:2016}.
One promising early attempt at accounting for correlated noise was the wavelet
approach of \citet{Carter:2009}.
Their technique allowed for a power-law spectrum for the noise scaling as
$\omega^{-\alpha}$, which unfortunately is not a normalizable, nor
physically-accurate, description of stellar variability.
Their approach runs in ${\cal O}(N)$ time for $\alpha=1$, which is referred to
as `flicker', $1/f$, or pink noise;  however, this particular type of noise
overpredicts the power spectrum at low frequency compared to the Solar
granulation power spectrum.  It also cannot describe quasi-periodic noise.

@@

The \celerite\ formalism also allows a principled approach to the estimation
of the power spectrum of a time series, as well as its uncertainties.
Fourier transforms work for uniformly sampled data;  astronomical data are
rarely (if ever) uniformly sampled.  Periodograms are affected by the window
function and aliasing in datasets, and make estimating the significance of
spectral features ambiguous.  The Bayesian approach to the estimation of a power
spectrum (CITE) avoids these difficulties:  it allows the computation of
uncertainties, it accounts for the window function on the uncertainties of
the power spectrum, and in principle it can treat aliases if the full posterior
is explored.  One drawback is that the number of $J$ terms to include is
unknown, but may be inferred with computation of the evidence, or an
approximation such as the Bayesian information criterion \citep{Kelly:2014}.
In general, the parameters of the \celerite\ terms are more interesting due
to their physical meaning rather than the overall shape of the power spectrum,
as shown in the foregoing examples.

@@

For arbitrary $J$, a procedure based upon Sturm's theorem \citep{Dorrie:1965}
allows one to determine whether there are any real roots within the range
$(0,\,\infty]$.
We first construct $Q_0(z)$ and its derivative $Q_1(z) = {Q_0}^\prime(z)$,
and then loop from $k=2$ to $k=2\,(J-1)+1$, computing
\begin{eqnarray}
Q_k(z) = -{\rm rem}(Q_{k-2},\,Q_{k-1})
\end{eqnarray}
where the function ${\rm rem}(p,\,q)$ is the remainder polynomial after
dividing $p(z)$ by $q(z)$.
